---
title: "Final_Assignment_part1"
author: "L.W"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)

```

This report will be looking at the historical shooting incident data from NYPD. We will be getting the data from this website <https://catalog.data.gov/dataset>. The file name will be "NYPD Shooting Incident Data (Historic)".

It contains more exploratory analysis than presented in the presentation.

Loading the needed libraries.
``` {r loadLibraries,eval=FALSE}
library(tidyverse)
library(lubridate)
```

## Step 1 - Import data

Per the data information page for this file ( <https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic>), the data contains every shooting incident that occurred in NYC going back to 2006 through the end of the previous calendar year (i.e. 2020).

Below is the R code to import the data and to show what data fields (and the corresponding data type) are in it. 
```{r dataImport, message=FALSE}

url<-"https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"

NY_shooting<-read_csv(url)
#spec(NY_shooting)

```

## Step 2 - Summary and clean up data

A quick examination of the data reveals that each record represents a shooting incident (uniquely identified by the `INCIDENT_KEY`) in NYC and includes information about the event, such as the location and time of occurrence, information related to suspect and victim demographics, and whether the event was flagged as a murder case. In total, there are `r nrow(NY_shooting)` records (incidences) with `r ncol(NY_shooting)` data fields.

The data is summarized here.  
``` {r SummaryRaw}
summary(NY_shooting)
```

We see that `OCCUR_DATE` is coded as character, we need to convert it into date. `PRECINCT` and `JURISDICTION_CODE` are indeed codes and should be changed into factor type, not as numbers. The two cases with unknown `JURISDICTION_CODE` will be left as NA if used. The variables for geographical locations will be removed since we don't need them for this analysis. At this time, we will also remove `JURISDICTION_CODE` since we don't plan to use it for this analysis, because we are not sure what this data represents. 

``` {r Clean}
NY_shooting_clean<-NY_shooting %>% select(-c(JURISDICTION_CODE,X_COORD_CD:Lon_Lat)) %>%
  mutate(eventDate = mdy(OCCUR_DATE), precinct = factor(PRECINCT))
#NY_shooting_clean %>% glimpse()
```

Checking the categorical variables for missing.
``` {r cateCheck}
missingCnt<-NY_shooting_clean %>% 
  summarize(across(LOCATION_DESC:VIC_RACE,~mean(is.na(.x))))
missingCnt
```

It seems `LOCATION_DESC`,  `PERP_AGE_GROUP`, `PERP_SEX`, `PERP_RACE` have `r paste0(round(min(missingCnt[missingCnt>0])/nrow(NY_shooting_clean)*100,1),'%')` or more missing. Decide to not use them for analysis either (i.e. remove). We'll also transform the `VIC_RACE` into a new field named `Race`, keeping "BLACK" and "WHITE" and mapping the other values as specified below.
``` {r detleteSevereMissingFields}
NY_shooting_clean<-NY_shooting_clean %>%
  select(-c(LOCATION_DESC,PERP_AGE_GROUP,PERP_SEX,PERP_RACE)) %>%
  mutate(Race=recode(VIC_RACE,
                     "BLACK HISPANIC" = "HISPANIC",
                     "WHITE HISPANIC" ="HISPANIC",
                     "ASIAN / PACIFIC ISLANDER"="Other",
                     "AMERICAN INDIAN/ALASKAN NATIVE"="Other"))

```

Another quick summary to take a look at the cleaned up data.
``` {r SummaryClean}
summary(NY_shooting_clean)
```

## Step 3 - Visualization and some analysis

#### Shooting incident by Boro

Let's take a look at the incident count for each BORO, and see which area has the most shooting incidents. 

``` {r BoroOverview}

metricByBoro<-NY_shooting_clean %>% 
  group_by(BORO) %>%
  summarize(murderPct = mean(STATISTICAL_MURDER_FLAG)*100, Incident_count = n())

ggplot(metricByBoro,aes(y = Incident_count, x = reorder(BORO,-Incident_count))) + #
  geom_col(aes(fill=reorder(BORO,-Incident_count)),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Incident Counts", 
       title="Shooting Incident Volume by BORO") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) 

  
```

Based on the bar graphs above, Brooklyn showed the most shooting incidents, followed by Bronx. 

Does it mean Brooklyn is the most unsafe place in NYC for everyone? We decide to take a look at the shooting volume by the victim's race and age group (ignoring the very small numbers of "UNKNOWN" cases in both fields).

``` {r BoroSubpop, message=FALSE}

by_boro_age_race<-NY_shooting_clean %>%
  filter(VIC_AGE_GROUP != "UNKNOWN" & VIC_RACE != "UNKNOWN") %>% # small number of unknown, ignore
  group_by(BORO,VIC_AGE_GROUP,Race) %>%
  summarize(incidentCnt=n()) %>% 
  mutate (pct = incidentCnt/sum(incidentCnt)*100)

by_boro_age_race %>% ungroup() %>% mutate(boroN =reorder(BORO,-incidentCnt,sum)) %>%
  ggplot(aes(y = incidentCnt, x = boroN)) +
  geom_col(aes(fill=boroN),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Incident Counts",
       title="Shooting Incident Volume by BORO",
       subtitle = "Displayed for Victim Age/Race Combination") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) +
  facet_grid(cols=vars(VIC_AGE_GROUP),rows=vars(Race))

```

Not surprisingly, there rarely were shooting victims of age 65 years old or older, regardless the race.

And, the shooting victims were overwhelmingly black or Hispanic. For White and "Other" races, there were much lower number of shooting victims. Because it was too small to see the counts in this graph for race "Other" and "WHITE", we zoomed the y-axis in for them by displaying only these two levels in the `Race` (below).

``` {r BoroSubPopFiltered}
by_boro_age_race %>% ungroup() %>% mutate(boroN =reorder(BORO,-incidentCnt,sum)) %>%
  filter(Race %in% c("WHITE","Other")) %>%
  ggplot(aes(y = incidentCnt, x = boroN)) +
  geom_col(aes(fill=boroN),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Incident Counts",
       title="Shooting Incident Volume by BORO",
       subtitle = "Displayed for Victim Age/Race Combination (Other and WHITE only)") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) +
  facet_grid(cols=vars(VIC_AGE_GROUP),rows=vars(Race))

```

In addition, we could look at the shooting incident volume for 2020 with and without normalization to the population of each Boro in the same year. The 2020 population info was obtained from this wiki page <https://en.wikipedia.org/wiki/Boroughs_of_New_York_City>. We could see that once taking into account the population in each Boro, the shooting rate is highest in Bronx instead of Brooklyn.

```{r boro2020Vol, message=FALSE}
metricByBoro2020<-NY_shooting_clean %>% 
  mutate(yr =  year(eventDate)) %>% 
  filter(yr == 2020) %>%
  group_by(BORO) %>%
  summarize(Incident_count = n()) %>% 
    ungroup %>%
    mutate(pop = c(1472654, 2736074, 1694251, 2405464, 495747)/1000000, Incident_rate = Incident_count/pop)
    
ggplot(metricByBoro2020,aes(y = Incident_count, x = reorder(BORO,-Incident_count))) + #
  geom_col(aes(fill=reorder(BORO,-Incident_count)),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Incident Counts", 
       title="2020 Shooting Incident Volume by BORO") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) 

ggplot(metricByBoro2020,aes(y = Incident_rate, x = reorder(BORO,-Incident_count))) + #
  geom_col(aes(fill=reorder(BORO,-Incident_count)),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="# of Incidents/million population", 
       title="2020 Shooting Incident Rate by BORO") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) 

  
```

The victim race composition of shooting incidents (below), stratified by the victim's age and the shooting location (Boro), showed that Black were more likely to be the victim in a shooting even in Boros with very low shooting incidents, such as Staten Island. 

It is interesting to note that in Staten Island, for age groups of 45-64 and 65+, the chance of White becoming a shooting victim increased dramatically. The proportion of White victims was almost neck-to-neck with that of the Black victims. It is intriguing to further investigate what might be the explanation behind this increase.

``` {r racePct}
by_boro_age_race %>% ungroup() %>% mutate(boroN =reorder(BORO,-incidentCnt,sum)) %>%
  ggplot(aes(y = pct, x = Race)) +
  geom_col(aes(fill=Race),alpha=0.5, show.legend=FALSE) +
  labs(x='Race', y="% of cases",
       title="Shooting Incident Victim Race Composition within each BORO & Age Combination") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) +
  facet_grid(cols=vars(VIC_AGE_GROUP),rows=vars(boroN))
```

#### Severity of the shooting incidents by Boro

Using the murder flag as a surrogate for the severity of the shooting incidents, display the proportion of murder incidents among the shooting events by Boro. The proportion of murder incidents among the shooting volume were not too different across the BOROs.

``` {r murderPct}

ggplot(metricByBoro,aes(y = murderPct, x = reorder(BORO,-Incident_count))) +
  geom_col(aes(fill=reorder(BORO,-Incident_count)),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Murder Incident (%)", 
       title="Percent Murder Shooting Incident by BORO") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8))

```

A logistic regression model on the BORO for the outcome of whether a shooting was flagged as murder was performed and confirmed that not enough evidence to indicate the severity of the shooting was different among the BOROs.

```{r logitMdl}
mdl<-glm(STATISTICAL_MURDER_FLAG~BORO, family=binomial, data = NY_shooting_clean)
summary(mdl)
anova(mdl,test = "LRT")
  
```

#### Shooting incident volume trend over time

We are also interested in seeing how the shooting volume changed over time. Let's transform the data and calculate the shooting volume by year, and visualize it. 

``` {r byYear}
by_boro_by_yr_cnt<-NY_shooting_clean %>% 
  mutate(yr = year(eventDate)) %>%
  group_by(BORO,yr) %>%
  summarize(cnt = n()) %>% 
  ungroup()

by_boro_by_yr_cnt %>%
  group_by(yr) %>%
  summarize(yrVol=sum(cnt)) %>%
  ggplot()+
  geom_point(aes(x=yr,y=yrVol),col='blue') +
  labs(x='Year', y="Number of shooting incidents", 
       title="Annual Shooting Incident Volume Over Time")

by_boro_by_yr_cnt %>%
  group_by(yr) %>%
  summarize(yrVol=sum(cnt))
```

The shooting incidents hovered close to 1900 for the first 6 years (2006-2011). It began a decreasing trend starting in 2012 and, within 6 years, reached a new stabilized level of below 1000, an almost 50% drop! However, in 2020, it bounced right back to the original high level, when the Covid-19 pandemic hit. 

When drilled down into the year trend in each Boro, prior to 2020, we see that Brooklyn had the largest improvement followed by Bronx. Queens and Manhattan were also in the direction of improvement, though at smaller magnitude. All 4 of these Boros had an up-tick of shooting incidents in 2020 that wiped out the previous years improvement. Whereas Staten Island maintained the same level of shooting incidents through out the 15 years.

``` {r byYearPrediction}
#fit <- lm(cnt~yr+BORO,data=by_boro_by_yr_cnt)
#summary(fit)
#by_boro_by_yr_cnt<- by_boro_by_yr_cnt %>%
 # mutate(pred = predict(fit))
by_boro_by_yr_cnt %>% mutate(boroN = reorder(BORO,-cnt,sum)) %>% 
  ggplot() +
  geom_point(aes(x=yr,y=cnt),col='blue') +
  #geom_point(aes(x=yr, y=pred),col='red') +
  facet_wrap(vars(boroN)) + 
  labs(x='Year', y="Number of shooting incidents", 
       title="Annual Shooting Incident Volume Over Time by BORO")
  
```


## Step 4 - Conslusion and Bias Identification

In conclusion, Brooklyn had the highest number of shooting incidents in years 2006-2020. If breaking down the shooting incident volume by victim age and race, we saw that most shooting involved BLACK and HIPANIC victims. Concerning the Black falling victims of shooting incidents, Brooklyn was the most dangerous neighborhood. However, concerning the Hispanic falling victims of shooting incidents, Bronx was the most dangerous neighborhood.

For the White and the other races, the numbers of shooting victims over the 15-year span were pretty low for each of the age groups. 

Based on the victim race composition stratified by Boro and age, Black consistently had the highest chance to be a shooting victim, even in area with very low shooting incidents, such as Staten Island. White usually had a low chance of being a shooting victim, except for those in the 45-64 and 65+ age groups from Staten Island. More data are likely needed to investigate the reasons behind this increase in the chance of White being shooting victims in such population. 

However, it is possible that higher number of victims of shooting for a certain age/race combination in a neighborhood might simply reflect the higher population of such demographic group in the area. To have a better understanding whether a certain age/race demographic are more likely to fall victims of shooting, we should normalize the shooting volume to the population in each area for each age/race combination. In current data, we unfortunately don't have the population information. 
In addition, we could also consider drilling down on the time of the day for the shooting incidents to learn if there is a pattern. This could help scheduling police patrol to hopefully prevent some of the shootings in the future.

The good news is that the annual shooting volume was overall in decline, though there was an up-tick in year 2020, likely related to the COVID pandemic in some ways and worth further exploration. 

Because we don't have details of how the data was collected, we need to be mindful of potential bias in the the data. For example, what might be causing the large proportion of perpetrators' demographic information to be missing? Simply ignoring those cases could bias the analysis. Hence this analysis chose not to use those variables. The number of UNKNOWN cases for the victims are fairly small (<0.5% of the data), the conclusions would be less susceptible to missing, though we should still emphasize the conclusion depends on the accuracy of the reporting of these data.


#### R session Info for the analysis is listed below

```{r sessionInfo, echo=FALSE}
sessionInfo()
```

```{r exploring,echo=FALSE,eval=FALSE}

NY_shooting_clean %>%
  mutate(BORO = BORO %>% factor() %>% fct_infreq() %>% fct_rev()) %>%
  ggplot(aes(BORO)) +
  geom_bar()

by_boro_age <- NY_shooting_clean %>%
  filter(VIC_AGE_GROUP != "UNKNOWN") %>% # small number, not interested
  count(BORO, VIC_AGE_GROUP) %>%
  group_by(BORO) %>%
  mutate(prop = n / sum(n)*100)


ggplot(by_boro_age,aes(y = prop, x = BORO)) +
  geom_col(aes(fill=BORO),alpha=0.5, show.legend=FALSE) +
  labs(x='BORO', y="Percent Shooting Incidence w/in BORO (%)", 
       title="Percent Shooting Incidence by Victim Age w/in BORO") +
  theme(axis.text.x=element_text(angle=90,hjust=1,size=8)) +
  facet_wrap(vars(VIC_AGE_GROUP))

ggplot(NY_shooting_clean,aes(BORO,fill=VIC_AGE_GROUP)) +geom_bar(position="fill")

ggplot(NY_shooting_clean,aes(BORO,fill=VIC_RACE)) +geom_bar(position="fill")

NY_shooting_clean %>% count(VIC_AGE_GROUP)

ggplot(NY_shooting_clean,aes(VIC_AGE_GROUP)) +geom_bar()

```

